---
title: "Simulate the SEIRV model"
author: "Sinead E. Morris"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Contact info

Github: <https://github.com/SineadMorris/trainings>

Email: [run7\@cdc.gov](mailto:run7@cdc.gov){.email}

Office: 7220.3 (Building 24, Floor 7)

### Load (and install) required packages

```{r load-packages}
# deSolve for solving model equations
# function used: 'ode()' 
# alternative functions: 'ode45()', 'lsoda()', 'lsode()'

# install to local R
if (! "deSolve" %in% installed.packages()){
  install.packages("deSolve")
}

# load into current environment
library(deSolve)
```

### Define model equations

We define the model equations inside a function.

Input:

-   t = time steps to simulate over
-   y vector = initial values for variables S, E, I, R, Sv, Ev, Iv, Rv
-   parms = vector of parameter values

Output:

-   Values of S, E, I, R, V at each time step

Note:

-   Since we now have two compartments in which individuals are capable of infecting others (I and Iv), we must incorporate both of these into the transmission term e.g. $\beta S I$ becomes $\beta S (I + I_v)$.
-   We must also make assumptions about the effectiveness of the vaccine.
    -   For instance, if the vaccine reduces the risk of infection by some fraction $f$, then we must reduce **transmission to the susceptible vaccinated** population by $(1 - f)$ i.e. $\beta (1 - f)  S_v (I + I_v)$.
    -   We could also assume that vaccination reduces the risk of onward transmission by some fraction $f_t$. In that case we would reduce **transmission from the infectious vaccinated** population by $(1 - f_t)$, i.e. $\beta S_v (I + (1 - f_t) I_v)$

```{r define-eqns}
eqns <- function(t, y, parms) {
  
  # use the names given in y (for the variables) & parms (for the parameters)
   with( as.list(c(y, parms)), {
    
    # Define change in S
    dS <- - beta * S * (I + Iv) - v * S
    
    # Define change in E
    dE <- beta * S * (I + Iv) - sigma * E
    
    # Define change in I
    dI <- sigma * E - gamma * I
    
    # Define change in R
    dR <- gamma * I
    
    # Define change in Sv
    dSv <- v * S - beta * (1 - f) * Sv * (I + Iv)
    
    # Define change in Ev
    dEv <- beta * (1 - f) * Sv * (I + Iv) - sigma * Ev
    
    # Define change in Iv
    dIv <- sigma * Ev - gamma * Iv
    
    # Define change in Rv
    dRv <- gamma * Iv
    
    # Keep track of values at each time step
     return(list( c(dS, dE, dI, dR, dSv, dEv, dIv, dRv) ))
  })
}
```

### Set parameter values

```{r set-parameters}
# define parameters we know
N  <- 10000
R0 <- 2
infectious_period <- 3
latent_period     <- 2

# v = vaccination rate per day, f = vaccine protection against infection
v <- 0.01
f <- 0.4

# Use these to get model parameters
gamma <- 1/infectious_period
sigma <- 1/latent_period

# Note: R0 = beta N / gamma  <=>  beta = R0 * gamma / N

parameters <- c(beta = R0 * gamma / N, gamma = gamma, sigma = sigma, v = v, f = f)

# define starting values ('initial conditions') for S, E, I, R, Sv, Ev, Iv, Rv
ics <- c(S = N - 1, E = 0, I = 1, R = 0, Sv = 0, Ev = 0, Iv = 0, Rv = 0)

# define time-steps to consider
days <- 1:365
```

### Simulate model

```{r simulate-model}
sim <- ode(y = ics, times = days, func = eqns, parms = parameters)
```

### Plot output

Note: we will just plot the output for the infected compartments (E, I, Ev, Iv) since plotting all compartments could get messy.

```{r plot-output}
# Create initial plot: t vs S
plot( x = sim[, "time"], y = sim[, "E"],              # x & y variables to plot
      type = "l",                                     # type of plot (l = line) 
      col = "black", lwd = 2.5,                       # color & width of line
      ylim = c(0, 500),                               # set limits of y-axis 
      xlab = "Days", ylab = "Number of people")       # add x & y axis labels

# Add lines for t vs I and t vs R
lines(x = sim[, "time"], y = sim[, "I"],  col = "darkmagenta", lwd = 2.5)
lines(x = sim[, "time"], y = sim[, "Ev"], col = "dodgerblue1", lwd = 2.5)
lines(x = sim[, "time"], y = sim[, "Iv"], col = "dodgerblue4", lwd = 2.5)

# Add legend to plot
legend(x = 300, y = 500,                               # position (x,y) = top-left corner                
       legend = c("E", "I", "Ev", "Iv"),               # names to display for variables
       col = c("black", "darkmagenta", "dodgerblue1", "dodgerblue4"),   
       pch = 16,                                       # symbol to display (16 = circle)
       cex = 1)                                        # size of legend
```

### Further resources

For this introduction we mainly used syntax and functions built-in to R (also referred to as 'base R'). Other packages we could have used include:

-   `dplyr` and `tidyr` for manipulating data frames
-   `ggplot2` for prettier plotting
-   `scico` and `viridis` for colorblind-friendly plotting palettes
-   `shiny`and `shinySIR` for making interactive apps
