---
title: "Simulate the SIR model"
author: "Sinead E. Morris"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Contact info

Github: https://github.com/SineadMorris/trainings

Email: run7@cdc.gov

Office: 7220.3 (Building 24, Floor 7)

### Load (and install) required packages

```{r load-packages}
# deSolve for solving model equations
# function used: 'ode()' 
# alternative functions: 'ode45()', 'lsoda()', 'lsode()'

# install to local R
if (! "deSolve" %in% installed.packages()){
  install.packages("deSolve")
}

# load into current environment
library(deSolve)
```


### Define model equations

We define the model equations inside a function.

Input:

* t = time steps to simulate over
* y vector = initial values for variables S, I, R
* parms = vector of parameter values

Output:

* Values of S, I, R at each time step

```{r define-eqns}
eqns <- function(t, y, parms) {
  
  # use the names given in y (for the variables) & parms (for the parameters)
   with( as.list(c(y, parms)), {
    
    # Define change in S
    dS <- - beta * S * I
    
    # Define change in I
    dI <- beta * S * I - gamma * I
    
    # Define change in R
    dR <- gamma * I
    
    # Keep track of values at each time step
     return(list( c(dS, dI, dR) ))
  })
}
```


### Set parameter values

```{r set-parameters}
# define parameters we know
N  <- 10000  
R0 <- 2
infectious_period <- 3

# Use these to get model parameters

# Note: infectious_period = 1 / gamma   <=>  gamma <- 1/ infectious_period
gamma <- 1/infectious_period 

# Note: R0 = beta N / gamma  <=>  beta = R0 * gamma / N

parameters <- c(beta = R0 * gamma / N, gamma = gamma)

# define starting values ('initial conditions') for S, I, R
ics <- c(S = N - 1, I = 1, R = 0)

# define time-steps to consider
days <- 1:150
```


### Simulate model

```{r simulate-model}
sim <- ode(y = ics, times = days, func = eqns, parms = parameters)
```


### Plot output

```{r plot-output}
# Create initial plot: t vs S
plot( x = sim[, "time"], y = sim[, "S"],               # x & y variables to plot
      type = "l",                                      # type of plot (l = line) 
      col = "black", lwd = 2.5,                        # color & width of line
      ylim = c(0, 10000),                               # set limits of y-axis 
      xlab = "Days", ylab = "Number of people")        # add x & y axis labels

# Add lines for t vs I and t vs R
lines(x = sim[, "time"], y = sim[, "I"], col = "dodgerblue1", lwd = 2.5)
lines(x = sim[, "time"], y = sim[, "R"], col = "dodgerblue4", lwd = 2.5)

# Add legend to plot
legend(x = 125, y = 7000,                              # position (x,y) = top-left corner                
       legend = c("S", "I", "R"),                      # names to display for variables
       col = c("black", "dodgerblue1", "dodgerblue4"), # colors to display
       pch = 16,                                       # symbol to display (16 = circle)
       cex = 1)                                        # size of legend
```


### Further resources

For this introduction we mainly used syntax and functions built-in to R (also referred to as 'base R'). Other packages we could have used include:

* `dplyr` and `tidyr` for manipulating data frames
* `ggplot2` for prettier plotting
* `scico` and `viridis` for colorblind-friendly plotting palettes
* `shiny`and `shinySIR` for making interactive apps



