---
title: "Simulating an SEIR model with births and deaths"
author: "James Mba Azam"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE
)
```

In this practical, we will be simulating an SIR model with a non-pharmaceutical intervention.

We will build on our previous SIR model code in `01_sir.Rmd` and modify it to include the time-varying transmission rate.

## The model function

We will start with the main model function.

```{r sir-npi-model}
# Define the SIR model differential equations
sir_npi <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    # Time-varying transmission rate
    beta <- ifelse(t < t_NPI_start | t > t_NPI_end, beta_0, (1 - epsilon) * beta_0)
    # SIR model
    dS <- -beta * S * I 
    dI <- beta * S * I - gamma * I
    dR <- gamma * I
    list(c(dS, dI, dR))
  })
}

```

Next, we will define the parameters and initial conditions. Note that we will also define the time at which the NPI is implemented, `t_NPI_start` and `t_NPI_end`.

## Initial conditions

```{r model-init}
# The initial conditions
# Population size
N  <- 1 
I_0 <- 0.01
R_0 <- 0
# Initial conditions for S, I, R
inits <- c(S = N - I_0, I = I_0, R = R_0)
```

## Model parameters

```{r model-params}
# Parameters
R0 <- 2
infectious_period <- 7
# Remember gamma = 1/ infectious_period as discussed earlier
gamma <- 1/infectious_period 
beta_0 <- 0.25
epsilon <- 0.5

# Combine the parameters
parms <- c(
  beta_0 = beta_0,
  gamma = gamma,
  epsilon = epsilon
)
```

## Time steps to return results

```{r model-time}
dt <- 1:365
```

## NPI implementation

```{r npi-time}
# Time at which the NPI is implemented
npi_times <- c(t_NPI_start = 10, t_NPI_end = 51) # 21 days of NPI  
```

------------------------------------------------------------------------

## Solving the model

Now, let's solve the model equations.

```{r sir-npi-solve}
# Load necessary packages
library(deSolve)
# Integrate the SIR equations over the time grid
output <- lsoda(
  y = inits,
  times = dt,
  func = sir_npi,
  parms = c(parms, npi_times)
)

# Convert the output to a data frame
output_df <- as.data.frame(output)
head(output_df)
```

------------------------------------------------------------------------

### Plotting the results

We will now plot the results by first converting the output to a long format and then using `ggplot2` to plot the data.

```{r sir-npi-plot}
# Load necessary packages
library(tidyr)
library(ggplot2)
# Convert the output to long format
output_df <- output_df |> pivot_longer(
  cols = -time,
  names_to = "compartment",
  values_to = "value"
)
# Plot the data using ggplot2
sir_npi_plot <- ggplot(
  data = output_df,
  aes(
    x = time,
    y = value,
    color = compartment
  )) +
  geom_line() +
  # Add a vertical line to indicate the times of NPI implementation
  geom_vline(
    xintercept = npi_times["t_NPI_start"],
    linetype = "dashed",
    color = "black"
  ) +
  geom_vline(
    xintercept = npi_times["t_NPI_end"],
    linetype = "dashed",
    color = "black"
  ) + 
  # Add labels
  labs(
    title = "SIR Model with NPI",
    x = "Time (days)",
    y = "Number of individuals",
    color = "Compartment"
  ) +
  theme_minimal()
plot(sir_npi_plot)
```

## Exercise

-   What do you observe when you change the value of `epsilon`?
-   What do you observe when you change the start and end times of the npi?
-   Would you recommend longer NPIs with smaller values of `epsilon` or shorter NPIs with larger values of `epsilon`?
-   How would you make this model more realistic?
